\vspace{-1.5em} \bf Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2

### 2023-06-08 paper.tex
Various updates

\subsection{The Li and Stephens model}
\label{sec:ls}
The Li and Stephens (LS) model \citep{Li2003-ib} is an approximation of the
coalescent with recombination~\citep{Hudson1983-properties}, which captures
many of the important features of the joint processes of mutation and
recombination. It is a Hidden Markov Model (HMM) in which a focal genome
is modelled as a
sequence of nucleotides that are probabilistically emitted as
an imperfect mosaic of a set of reference genomes
(Figure~\ref{fig:ls_diagram}).
The LS model is used in a wide variety
of applications in genomics, including modern approaches to
statistical genotype phasing and imputation
\citep{Delaneau2019-wl,Browning2021-cg,Browning2018-nk,Rubinacci2020-pa},
and estimation of parameters such as
recombination rates \citep[e.g.,][]{Hinch2011-tz}
and intensity of selection within and across hosts in viral
sequence data \citep[e.g.,][]{Palmer2019-wa}.
See~\cite{Mcvean2019-linkage} for further review and discussion
of the LS model.

@@

The generative process of the LS model is summarised in
Figure~\ref{fig:ls_diagram}. Here, a transition matrix, $Q$, governs the
process of switching (recombining) between members of the reference panel (the
hidden states). An emission matrix, $E$, allows for differences between the
focal sequence and the hidden state from which it is copied (due to mutation as
well as sequencing error).
Both $E$ and $Q$ may be a function of the reference panel members,
but transitions are generally assumed to be independent of the
hidden states (Figure~\ref{fig:ls_diagram}, pink panel).
This assumption dramatically increases performance as the state space drops to two states (i.e., switching or not switching).
Emissions may also be a function of the nucleotide states, but in our RNA virus case, we assume that \sally{mutations occur to all possible alternate alleles} ($A,C,G,U,-$) with equal probability for a mutation to any other allele (including a gap in the alignment; $-$) are equally likely $\mu_\ell/4$. This is reasonable for rapidly evolving pathogens, but we note that setting the number of alleles at site $\ell$ ($a_\ell$) to the set of observed alleles across all analysed samples ($\mu_\ell/(a_\ell-1)$) will often be more appropriate
 (Figure~\ref{fig:ls_diagram}, blue panel).
We use the Viterbi algorithm~\citep{Viterbi1967-ol}
to find the most likely copying path, given $Q$, $E$, and a set of reference sequences.
Throughout, we refer to the probabilities of mismatching to a
member of the reference panel at site $\ell$ as $\mu_\ell$,
and the probability of recombining between two members of the reference panel between site $\ell-1$ and site $\ell$ as $r_\ell$. For convenience, $\mu_\ell$ is commonly referred to as the `mutation probability', but we note that this probability of
mismatch also encompasses various other error modes that result in a mismatch
(such as sequencing and alignment errors). Note also that it is these probabilities,
not the rates of mutation and recombination, that are required to fully define
the HMM; see~\cite{Donnelly2010-coalescent} for a discussion of how these
parameters relate to the coalescent process.

@@

The probabilities of these competing processes of mismatch and recombination are
usually controlled by the site-specific parameters $\mu_\ell$ and $r_\ell$,
respectively. For this work we used a slightly different formulation, which
uses one parameter, the mismatch ratio (MMR), to control the relative
importance of mutation and recombination in the HMM. Specifically, an MMR value
of $k$ will prefer $k$ mismatches (mutations) to a single recombination that
 results in copying from a template with no mismatches. To map between
recombination probability and mutation probability for a particular mismatch
ratio, we simply consider the two paths that we wish to be equally likely and
rearrange for the mutation or recombination probability. Consider the simple
case where we assume $\mu_l=\mu$ and $r_l=r$. Without loss of generality,
consider a region of length $m$. Up until this region, two paths are equally
likely, so we can re-scale by the likelihood of observing the focal sequence up
to the site before the region starts, $c$. For an MMR of $k$, we only need to
consider two of the $n$ members of the reference panel.
One, $\mathcal{P}_1$,
for which there are no mismatches, but we need to recombine in, and a second,
$\mathcal{P}_2$, for which there are $k$ (randomly chosen) mismatches in the
region.
Now, follow the probability of tracking along each of those paths:
\begin{align*}
\mathbb{P}[\mathcal{P}_1] &= \frac{cr}{n}
\alpha^{m-1}\left(1-\mu\right)^m
&\text{recombine to a template with no mismatches,}\\
\mathbb{P}[\mathcal{P}_2] &= c
\alpha^{m}
% \alpha \alpha^{m-1}
\left(1-\mu\right)^{m-k}\mu^k &\text{stay in a template with $k$ mismatches,}
\end{align*}
% Note: brackets here emphasise the two different components 
% of alpha: (1-r) = don't leave this reference member, 
% and r/n = move out, but back to yourself.
where $\alpha = (1 - r) + r/n$ represents the probability of a sequence not recombining at a given position with any of the other $n-1$ members of the reference panel (Figure~\ref{fig:ls_diagram}, red panel) and $c$ is the
likelihood of the path up to this point (assumed to be
equal by construction).
We then
set these path probabilities to be equal, and rearrange to relate $r$ and $\mu$ to one another:
\begin{align*} r = \frac{n\mu^k}{\left(1-\mu\right)^k + \left(n-1\right)\mu^k},
\quad\quad \mu = \frac{1}{\sqrt[k]{\frac{n}{r} - (n-1)} + 1}. \end{align*}
Thus, for lower MMR values ($k$ here),
recombination is increasingly favoured over multiple mutations in a specific ancestral genome.
We use an MMR value of $k=3$ in this work,
because of the relatively high rate of recombination relative 
to mutation typical of 
coronaviruses~\citep{amoutzias2022remarkable}.
Exploring different mismatch ratios and more sophisticated parameterisations
of the HMM are important avenues for future work.

@@

\begin{figure} \centering
\includegraphics[width=0.85\textwidth]{figures/false_positive_top2_nxcld_large_graph.pdf}
\caption{\label{fig:false_positives}
Subgraph of the Long ARG, focussing on two likely false positive 
recombination nodes at the start of the Delta wave 
(``focal'' nodes, in red, corresponding to rows A and B in 
Table~\ref{tab:false_positive}).
The path to node \texttt{tsk261771} has been expanded: 
this node (in gold) is a large polytomy with
107 immediate children, and is the ancestor of the majority of Delta samples.
The suspected-incorrect recombination path to an early B.1 MRCA is also shown.
Note the large amount of additional recombination (black nodes) among close 
descendants of the focal nodes.
% Why say this here and not at for the other subgraph figs?
%See supplementary file \protect\path{Subgraph_sample_mapping.txt} to  map tskID to EPI\_ISL and strain name.
}
\end{figure}

>>>>

\subsection{The Li and Stephens model}
\label{sec:ls}
The Li and Stephens (LS) model \citep{Li2003-ib} is an approximation of the
coalescent with recombination~\citep{Hudson1983-properties} which captures
many of the important features of the joint processes of mutation and
recombination. It is a Hidden Markov Model (HMM) in which a focal genome
is modelled as a
sequence of nucleotides that are probabilistically emitted as
an imperfect mosaic of a set of reference genomes
(Figure~\ref{fig:ls_diagram}).
The LS model is used in a wide variety
of applications in genomics, including modern approaches to
statistical genotype phasing and imputation
\citep{Delaneau2019-wl,Browning2021-cg,Browning2018-nk,Rubinacci2020-pa},
and estimation of parameters such as
recombination rates \citep[e.g.,][]{Hinch2011-tz}
and intensity of selection within and across hosts in viral
sequence data \citep[e.g.,][]{Palmer2019-wa}.
See~\cite{Mcvean2019-linkage} for further review and discussion
of the LS model.

@@

The generative process of the LS model is summarised in
Figure~\ref{fig:ls_diagram}. Here, a transition matrix, $Q$, governs the
process of switching (recombining) between members of the reference panel (the
hidden states). An emission matrix, $E$, allows for differences between the
focal sequence and the hidden state from which it is copied (due to mutation as
well as sequencing error).
Both $E$ and $Q$ may be a function of the reference panel members,
but transitions are generally assumed to be independent of the
hidden states (Figure~\ref{fig:ls_diagram}, pink panel).
This assumption dramatically increases performance as the state space drops to two states (i.e., switching or not switching).
Emissions may also be a function of the nucleotide states, 
but in our RNA virus case
we assume that mutations occur from all possible alleles 
($A,C,G,U$ and a gap in the alignment, $-$) to any other with
equal probability $\mu_\ell/4$.
This is reasonable for rapidly evolving pathogens, 
but we note that setting the number of alleles at site $\ell$ ($a_\ell$) 
to the set of observed alleles across all analysed 
samples ($\mu_\ell/(a_\ell-1)$) will often be more appropriate 
(Figure~\ref{fig:ls_diagram}, blue panel).
We use the Viterbi algorithm~\citep{Viterbi1967-ol}
to find the most likely copying path, given $Q$, $E$, and a set of reference sequences.
Throughout, we refer to the probabilities of mismatching to a
member of the reference panel at site $\ell$ as $\mu_\ell$,
and the probability of recombining between two members of the reference panel between site $\ell-1$ and site $\ell$ as $r_\ell$. For convenience, $\mu_\ell$ is commonly referred to as the `mutation probability', but we note that this probability of
mismatch also encompasses various other error modes that result in a mismatch
(such as sequencing and alignment errors). Note also that it is these probabilities,
not the rates of mutation and recombination, that are required to fully define
the HMM; see~\cite{Donnelly2010-coalescent} for a discussion of how these
parameters relate to the coalescent process.

@@

The probabilities of these competing processes of mismatch and recombination are
usually controlled by the site-specific parameters $\mu_\ell$ and $r_\ell$,
respectively. For this work we used a slightly different formulation, which
uses one parameter, the mismatch ratio (MMR), to control the relative
importance of mutation and recombination in the HMM. Specifically, an MMR value
of $k$ will prefer $k$ mismatches (mutations) to a single recombination that
 results in copying from a template with no mismatches. To map between
recombination probability and mutation probability for a particular mismatch
ratio, we simply consider the two paths that we wish to be equally likely and
rearrange for the mutation or recombination probability. Consider the simple
case where we assume $\mu_l=\mu$ and $r_l=r$. Without loss of generality,
consider a region of length $m$. Up until this region, two paths are equally
likely, so we can re-scale by the likelihood of observing the focal sequence up
to the site before the region starts, $c$. For an MMR of $k$, we only need to
consider two of the $n$ members of the reference panel.
One, $\mathcal{P}_1$,
for which there are no mismatches, but we need to recombine in, and a second,
$\mathcal{P}_2$, for which there are $k$ (randomly chosen) mismatches in the
region.
The probability of tracking along each of those paths is given by:
\begin{align*}
\mathbb{P}[\mathcal{P}_1] &= \frac{cr}{n}
\alpha^{m-1}\left(1-\mu\right)^m
&\text{recombine to a template with no mismatches,}\\
\mathbb{P}[\mathcal{P}_2] &= c
\alpha^{m}
% \alpha \alpha^{m-1}
\left(1-\mu\right)^{m-k}\mu^k &\text{stay in a template with $k$ mismatches,}
\end{align*}
% Note: brackets here emphasise the two different components 
% of alpha: (1-r) = don't leave this reference member, 
% and r/n = move out, but back to yourself.
where $\alpha = (1 - r) + r/n$ represents the probability of a sequence not recombining at a given position with any of the other $n-1$ members of the reference panel (Figure~\ref{fig:ls_diagram}, red panel) and $c$ is the
likelihood of the path up to this point (assumed to be
equal by construction).
We then
set these path probabilities to be equal, and rearrange to relate $r$ and $\mu$ to one another:
\begin{align*} r = \frac{n\mu^k}{\left(1-\mu\right)^k + \left(n-1\right)\mu^k},
\quad\quad \mu = \frac{1}{\sqrt[k]{\frac{n}{r} - (n-1)} + 1}. \end{align*}
Thus, for lower MMR values ($k$ here),
recombination is increasingly favoured over multiple mutations in a specific ancestral genome.
We use an MMR value of $k=3$ in this work,
because of the relatively high rate of recombination relative 
to mutation typical of 
coronaviruses~\citep{amoutzias2022remarkable}.
Exploring different mismatch ratios and more sophisticated parameterisations
of the HMM are important avenues for future work.

@@

\begin{figure} \centering
\includegraphics[width=0.85\textwidth]{figures/false_positive_top2_nxcld_large_graph.pdf}
\caption{\label{fig:false_positives}
Subgraph of the Long ARG, focussing on two likely false positive 
recombination nodes at the start of the Delta wave 
(``focal'' nodes, in red, corresponding to rows A and B in 
Table~\ref{tab:false_positive}).
The path to node \texttt{tsk261771} has been expanded: 
this node (in gold) is the ancestor of ${\sim}90\%$ of Delta samples in the Long ARG,
and represents a large polytomy with 107 immediate children.
The suspected-incorrect recombination path to an early B.1 MRCA is also shown.
Note the large amount of additional recombination (black nodes) among close 
descendants of the focal nodes.
% Why say this here and not at for the other subgraph figs?
%See supplementary file \protect\path{Subgraph_sample_mapping.txt} to  map tskID to EPI\_ISL and strain name.
}
\end{figure}
